---
title: "Blois et al 2014 test"
author: "D Li"
date: "11/20/2014"
output:
  html_document:
    keep_md: yes
---
```{r knitr_options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align='center') 
options(width = 90)
```

This project tests the effectiveness of Blois et al 2014 Ecography paper's framework to disentangle different mechanisms for co-occurrent species. It has the following steps:

1. Test the randomness of overall co-occurrence patterns. Results were stored in `cscores`.
2. Test the significance of co-occurrence for all possible species pairs. In the test dataset, we have 100 speciese, so we have `choose(100, 2) = 4950` pairs. We will collect all significant positive (`pairs.pos`) and negative (`pairs.neg`) co-occurrent species pairs.
3. Test the underlying mechanisms of each of these significant pairs. Particularly, we only tested *environmental filtering* and *species interactions* here, without testing *dispersal limitation*. For positive pairs, we compared sites that they both present and sites that they both absent. If these two groups of sites are significantly different in environmental conditions, then we infer the possible mechanism as environmental filtering. If not, we infer as species interaction.
  - However, they are some issues:
      + we may not include all important environmental varibles.
      + for the same pair, we may conclude that it was shaped by environmental filtering by one environmental variable and species interaction by another environmental variable.
4. Get the proportion of species pairs that can be explained by either mechanism for each environmental variable.

See R codes for details.

### Data
```{r data}
## data----
load("xy.Rdata")
envi = data.frame(x)
row.names(envi) = paste0("site", 1:2500)
colnames(envi) = paste0("envi", 1:5)
envi$lat = runif(n = 2500, min = 40.00000, max = 45.999999)
envi$long = runif(n = 2500, min = -90.00000, max = -80.999999)
envi$site = row.names(envi) 
envi = select(envi, site, long, lat, envi1:envi5)

veg = data.frame(y)
row.names(veg) = paste0("site", 1:2500)
colnames(veg) = paste0("sp", 1:100)
rm(x, y)

# sort(colSums(veg))
# 
# ## plot
par(mfrow=c(3,2))
hist(envi$envi1)
hist(envi$envi2)
hist(envi$envi3)
hist(envi$envi4)
hist(envi$envi5)
hist(rowSums(veg))
# hist(colSums(veg))
# par(mfrow=c(1,1))
# plot(envi$envi1, rowSums(veg))
```


```{r load results}
# load the results so can avoid re-run
load("testresult.RData")
```

### Overall co-occurrence pattern (site-level)
Then let's look at the overall randomness of co-occurrence patterns. C-score is the average of all checkboad units. Thus, higher cscores suggest stronger competition among species. But we cannot naively say that species competition is the main mechanisms. Because things like habitat preferences, dispersal ability, etc. can also get the same pattern.
```{r cscore, eval=FALSE}
cscores = cs_sim(veg, nullmethod = "tswap", n = 5000)
```
```{r cscore2}
cscores
```
It is highly significant, which suggests that the overall co-occurrence pattern is non-random. Potentially, species competition is very important.

### Pairwise co-occurrence patterns (species-level)
Then we will test the significance of co-occurrence for each species pair.
```{r pairs, eval=FALSE}
pairs.all = pair.ff(comm = veg, nullmethod = "tswap", formu = "(A-J)*(B-J)", nsim = 1000, burn = 10)
pairs.all = p.adj(pairs.all)

pairs.pos = filter(pairs.all, SES > 0 & p.value < 0.05)
pairs.neg = filter(pairs.all, SES < 0 & p.value < 0.05)

pairs.all.summary = data.frame(sp.rich = length(unique(c(pairs.all$sp1_name, pairs.all$sp2_name))),
            all.pos = sum(pairs.all$SES>0), all.neg = sum(pairs.all$SES<0), 
            sig.pos = sum(pairs.all$SES>0 & pairs.all$p.value < 0.05),
            sig.neg = sum(pairs.all$SES<0 & pairs.all$p.value < 0.05),
            all.pos.prop = sum(pairs.all$SES>0)/choose(length(unique(c(pairs.all$sp1_name, pairs.all$sp2_name))), 2), 
            all.neg.prop = sum(pairs.all$SES<0)/choose(length(unique(c(pairs.all$sp1_name, pairs.all$sp2_name))), 2), 
            sig.pos.prop = sum(pairs.all$SES>0 & pairs.all$p.value < 0.05)/choose(length(unique(c(pairs.all$sp1_name, pairs.all$sp2_name))), 2),
            sig.neg.prop = sum(pairs.all$SES<0 & pairs.all$p.value < 0.05)/choose(length(unique(c(pairs.all$sp1_name, pairs.all$sp2_name))), 2))
```
```{r pairs2}
head(pairs.all)
head(pairs.pos)
head(pairs.neg)
pairs.all.summary
```
I found 65% of species pairs to be significant, which is very high compare with real dataset.

### Get explanation for co-occurrence pairs
```{r envi,eval=FALSE}
#### envi diff of pairs ----
explain.pos = get_together(veg = veg, cooc = pairs.pos, envi = envi, pos.neg = "pos", m = 5)
explain.neg = get_together(veg = veg, cooc = pairs.neg, envi = envi, pos.neg = "neg", m = 5)
```
```{r envi2}
explain.all.summary = rbind(explain.pos[[1]], explain.neg[[1]])
names(explain.all.summary)[2] = "Explanation"
names(explain.all.summary)[3] = "Proportion"
explain.all.summary 

library(ggplot2)
ggplot(explain.all.summary, aes(x=variable, y=Proportion, fill = Explanation)) + 
  geom_bar(stat="identity") +
  facet_wrap(~posneg) +
  theme(legend.position="top", axis.title.x=element_blank()) +
  guides(fill = guide_legend(title = "Causal explanation"))
```
For positive co-occurrent pairs, about 70% of them were driven by environmental filtering. Other ~30% were shaped by potential interactions, e.g. facilitation, among species. For negative co-occurrent pairs, about 75% of them were driven by biotic competition. 

### How good is this framework?


